---
title: "M1 Per-Trait QC Report"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    embed-resources: true
    self-contained: true
params:
  trait: "bmi"
  ancestry: "EUR"
  consortium: "GIANT-UKBB"
  year: "2018"
  parquet: ""
  harmonized_tsv: ""
  rg_log_dir: "data/processed/ldsc_overlap/rg_logs"
  qc_json: ""
  control_loci_csv: "src/R/qc/control_loci.csv"
  sha256: ""
  build: "GRCh37"
execute:
  echo: false
  warning: false
  message: false
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(arrow)
  library(ggplot2)
  library(dplyr)
  library(qqman)
  library(jsonlite)
  library(knitr)
})

trait_key <- sprintf("%s.%s.%s.%s", params$trait, params$ancestry,
                     params$consortium, params$year)
```

# Per-Trait QC Report — `r trait_key`

* **Trait**: `r params$trait`
* **Ancestry**: `r params$ancestry`
* **Consortium**: `r params$consortium`
* **Year**: `r params$year`
* **Build**: `r params$build`
* **Parquet**: `r params$parquet`
* **SHA-256 (raw)**: `r params$sha256`

This report walks through the 9-item QC checklist defined in
`SUMSTATS-UPGRADE.md §7`. Each section emits PASS / WARN / FAIL.

```{r read_data}
parquet_exists <- file.exists(params$parquet)
df <- if (parquet_exists) {
  arrow::read_parquet(params$parquet)
} else {
  data.frame()
}
n_rows <- nrow(df)
```

## 1. Variant count

* **n_rows**: `r format(n_rows, big.mark = ",")`
* **Threshold**: ≥ 3,000,000 (D-12 sanity; HM3 munge will further restrict)
* **Status**: `r if (n_rows >= 3e6) "PASS" else if (n_rows >= 1e6) "WARN" else "FAIL"`

## 2. MAF distribution

```{r maf_hist}
maf_col <- intersect(c("EAF", "FRQ", "MAF", "AF1"), names(df))[1]
maf_status <- "SKIP"
maf_zero_pct <- NA_real_
if (!is.na(maf_col) && n_rows > 0) {
  af <- as.numeric(df[[maf_col]])
  maf <- pmin(af, 1 - af)
  maf <- maf[!is.na(maf)]
  maf_zero_pct <- mean(maf == 0) * 100
  maf_status <- if (maf_zero_pct < 5) "PASS" else "WARN"
  ggplot(data.frame(maf = maf), aes(x = maf)) +
    geom_histogram(bins = 50, fill = "steelblue", colour = "white") +
    theme_minimal() +
    labs(title = sprintf("MAF distribution (%s; n=%s)",
                         maf_col, format(length(maf), big.mark = ",")),
         x = "MAF", y = "count")
}
```

* **Status**: `r maf_status` (MAF=0 fraction = `r ifelse(is.na(maf_zero_pct), NA, sprintf("%.2f%%", maf_zero_pct))`)

## 3. Build verification

* **Plan-spec'd build**: `r params$build`
* **Status**: PASS (canonical D-01 GRCh37 enforced upstream by harmonizers)

## 4. Effect-allele labeling

```{r ea_check}
ea_col <- intersect(c("EA", "ALT", "A1"), names(df))[1]
oa_col <- intersect(c("OA", "REF", "A2"), names(df))[1]
ea_status <- if (!is.na(ea_col) && !is.na(oa_col)) "PASS" else "FAIL"
```

* **EA column**: `r ifelse(is.na(ea_col), "NOT FOUND", ea_col)`
* **OA column**: `r ifelse(is.na(oa_col), "NOT FOUND", oa_col)`
* **Status**: `r ea_status`

## 5. LDSC intercept and h²

```{r ldsc_parse}
parse_h2_from_log <- function(log_path, trait_key) {
  if (!file.exists(log_path)) return(list(h2_obs = NA_real_, h2_int = NA_real_,
                                          gcov_int = NA_real_))
  txt <- readLines(log_path)
  in_table <- FALSE
  for (line in txt) {
    if (grepl("Summary of Genetic Correlation Results", line)) {
      in_table <- TRUE
      next
    }
    if (!in_table) next
    parts <- strsplit(trimws(line), "\\s+")[[1]]
    if (length(parts) < 12) next
    p1 <- parts[1]; p2 <- parts[2]
    if (grepl(trait_key, p1) || grepl(trait_key, p2)) {
      return(list(h2_obs = suppressWarnings(as.numeric(parts[7])),
                  h2_int = suppressWarnings(as.numeric(parts[9])),
                  gcov_int = suppressWarnings(as.numeric(parts[11]))))
    }
  }
  list(h2_obs = NA_real_, h2_int = NA_real_, gcov_int = NA_real_)
}

ldsc <- list(h2_obs = NA_real_, h2_int = NA_real_)
if (dir.exists(params$rg_log_dir)) {
  logs <- list.files(params$rg_log_dir, pattern = "^focal_.*\\.log$",
                     full.names = TRUE)
  for (lg in logs) {
    res <- parse_h2_from_log(lg, trait_key)
    if (!is.na(res$h2_obs) || !is.na(res$h2_int)) {
      ldsc <- res
      break
    }
  }
}
ldsc_int_status <- if (is.na(ldsc$h2_int)) "SKIP" else if (ldsc$h2_int >= 0.9 && ldsc$h2_int <= 1.15) "PASS" else "WARN"
```

* **h²_obs**: `r ifelse(is.na(ldsc$h2_obs), NA, sprintf("%.4f", ldsc$h2_obs))`
* **h²_int**: `r ifelse(is.na(ldsc$h2_int), NA, sprintf("%.4f", ldsc$h2_int))`
* **Status (LDSC intercept in [0.9, 1.15])**: `r ldsc_int_status`

## 6. λ_GC

```{r lambda_gc}
p_col <- intersect(c("P", "PVAL", "p_value", "Pvalue"), names(df))[1]
lambda_gc <- NA_real_
if (!is.na(p_col) && n_rows > 0) {
  pvals <- suppressWarnings(as.numeric(df[[p_col]]))
  pvals <- pvals[!is.na(pvals) & pvals > 0 & pvals <= 1]
  if (length(pvals) > 1000) {
    chisq <- qchisq(1 - pvals, df = 1)
    lambda_gc <- median(chisq, na.rm = TRUE) / qchisq(0.5, df = 1)
  }
}
lambda_status <- if (is.na(lambda_gc)) "SKIP" else if (lambda_gc >= 0.9 && lambda_gc <= 1.15) "PASS" else if (lambda_gc <= 1.2) "WARN" else "FAIL"
```

* **λ_GC**: `r ifelse(is.na(lambda_gc), NA, sprintf("%.4f", lambda_gc))`
* **Status (in [0.9, 1.15])**: `r lambda_status`

## 7. Positive-control loci

```{r control_loci}
ctrl <- if (file.exists(params$control_loci_csv)) {
  read.csv(params$control_loci_csv)
} else data.frame()
trait_ctrl <- ctrl[ctrl$trait == params$trait, , drop = FALSE]
chr_col <- intersect(c("CHR", "chr"), names(df))[1]
bp_col <- intersect(c("BP", "POS", "bp"), names(df))[1]
present_status <- "SKIP"
hits_summary <- "no control loci registered for this trait"
if (nrow(trait_ctrl) > 0 && !is.na(chr_col) && !is.na(bp_col) && n_rows > 0) {
  hits <- list()
  for (i in seq_len(nrow(trait_ctrl))) {
    row <- trait_ctrl[i, ]
    win_lo <- as.numeric(row$pos_b37) - 5e5
    win_hi <- as.numeric(row$pos_b37) + 5e5
    sub <- df[df[[chr_col]] == as.character(row$chr) &
              df[[bp_col]] >= win_lo & df[[bp_col]] <= win_hi, , drop = FALSE]
    n_hit <- nrow(sub)
    hits[[length(hits) + 1]] <- data.frame(
      locus = row$locus_name,
      chr = row$chr,
      pos_b37 = row$pos_b37,
      n_variants_in_window = n_hit
    )
  }
  hits_df <- do.call(rbind, hits)
  hits_summary <- knitr::kable(hits_df)
  present_status <- if (all(hits_df$n_variants_in_window > 0)) "PASS" else "WARN"
}
```

* **Status**: `r present_status`

`r if (is.character(hits_summary)) hits_summary`

## 8. Strand-ambiguous (palindromic) drop rate

```{r palindromic}
palin_pct <- NA_real_
palin_status <- "SKIP"
if (file.exists(params$qc_json)) {
  qc <- jsonlite::fromJSON(params$qc_json)
  n_in <- qc$n_input
  n_palin <- qc$n_palindromic_dropped
  if (!is.null(n_in) && !is.null(n_palin) && is.numeric(n_in) && n_in > 0) {
    palin_pct <- 100 * n_palin / n_in
    palin_status <- if (palin_pct < 10) "PASS" else "FAIL"
  }
}
```

* **palindromic drop %**: `r ifelse(is.na(palin_pct), NA, sprintf("%.3f%%", palin_pct))`
* **Status (< 10%)**: `r palin_status`

## 9. PASS / FAIL summary

```{r summary_table}
summary_df <- data.frame(
  item = c("1. Variant count >= 3M",
           "2. MAF distribution (MAF=0 < 5%)",
           "3. Build = GRCh37",
           "4. EA / OA columns present",
           "5. LDSC intercept in [0.9, 1.15]",
           "6. λ_GC in [0.9, 1.15]",
           "7. Control-locus presence",
           "8. Palindromic drop < 10%",
           "9. Per-variant N integrity"),
  status = c(if (n_rows >= 3e6) "PASS" else if (n_rows >= 1e6) "WARN" else "FAIL",
             maf_status,
             "PASS",
             ea_status,
             ldsc_int_status,
             lambda_status,
             present_status,
             palin_status,
             "WARN-MANUAL")
)
knitr::kable(summary_df, caption = "9-item §7 QC checklist")
```

```{r emit_qc_status, include=FALSE}
all_pass <- all(summary_df$status %in% c("PASS", "SKIP", "WARN", "WARN-MANUAL"))
qc_status <- if (any(summary_df$status == "FAIL")) "FAIL" else "PASS"
```

**Overall trait QC status**: `r qc_status`
