---
title: "Phase 1 -- SuSiE fine-mapping QC dashboard"
subtitle: "REQ-2 D1+D2+D3+D4+D6"
format:
  html:
    embed-resources: true
    toc: true
params:
  input_tsv: "results/finemap/qc_aggregated.tsv"
  sweep_tsv: "results/finemap/sweep_complex_regions.tsv"
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(data.table)
  library(DT)
})
qc <- fread(params$input_tsv)
```

## Summary

```{r summary-stats, echo=FALSE}
cat("Total rows:", nrow(qc), "\n")
cat("Unique regions:", length(unique(qc$region_id)), "\n")
cat("Unique traits:", length(unique(qc$trait)), "\n")
cat("Unique ancestries:", length(unique(qc$ancestry)), "\n")
cat("Non-converged:", sum(qc$convergence_status == "non_converged", na.rm = TRUE), "\n")
cat("L-saturated:", sum(qc$L_saturated == TRUE, na.rm = TRUE), "\n")
cat("Block-diagonal LD (HLA):", sum(qc$ld_source == "ukbb_ld_tiled_block_diagonal", na.rm = TRUE), "\n")
cat("Complex regions:", sum(qc$is_complex_region == TRUE, na.rm = TRUE), "\n")
```

## Per-locus table

Row styling (applied via `DT::formatStyle` conditional formatting):

- **Red** background if `ld_source == ukbb_ld_tiled_block_diagonal` (HLA statistical caveat, T-1-04)
- **Orange** background if `convergence_status != converged_primary` (retry ladder exercised)
- **Yellow** background if `L_saturated` (all L=10 effects retained)

```{r per-locus-table, echo=FALSE}
DT::datatable(
  qc,
  options = list(
    pageLength = 50,
    deferRender = TRUE,
    scrollX = TRUE,
    autoWidth = TRUE
  ),
  filter = "top",
  rownames = FALSE,
  class = "display compact"
) |>
  DT::formatStyle(
    "ld_source",
    backgroundColor = DT::styleEqual(
      c("ukbb_ld_tiled_block_diagonal"),
      c("#ffcccc")
    )
  ) |>
  DT::formatStyle(
    "convergence_status",
    backgroundColor = DT::styleEqual(
      c("non_converged", "converged_regularized", "converged_max_iter"),
      c("#ff9999", "#ffd699", "#fff2cc")
    )
  ) |>
  DT::formatStyle(
    "L_saturated",
    backgroundColor = DT::styleEqual(c("TRUE", "True", "true"), rep("#ffff99", 3))
  ) |>
  DT::formatStyle(
    "is_complex_region",
    backgroundColor = DT::styleEqual(c("TRUE", "True", "true"), rep("#e6ccff", 3))
  )
```

## REQ-2 Complex-region sensitivity sweep

```{r sweep-table, echo=FALSE}
sweep_path <- params$sweep_tsv
if (file.exists(sweep_path)) {
  # The sweep TSV may have a comment header; skip lines starting with ##
  sweep_raw <- readLines(sweep_path)
  sweep_data <- sweep_raw[!grepl("^##", sweep_raw)]
  if (length(sweep_data) > 1) {
    sweep <- fread(text = paste(sweep_data, collapse = "\n"))
    DT::datatable(
      sweep,
      options = list(
        pageLength = 50,
        deferRender = TRUE,
        scrollX = TRUE
      ),
      filter = "top",
      rownames = FALSE,
      caption = "Supplementary: sensitivity sweep across min_abs_corr thresholds for complex regions"
    )
  } else {
    cat("Sweep table is empty (no known_complex or data_flagged rows on current data).\n")
  }
} else {
  cat("Sweep table not yet generated.\n")
}
```

## Notes

- `ld_source == "ukbb_ld_tiled_block_diagonal"` is the HLA_6p21 block-diagonal approximation (T-1-04 surfacing).
  Cross-tile off-diagonal LD blocks are set to zero -- interpret HLA CS purity as optimistic.
- `convergence_status != "converged_primary"` indicates the retry ladder was exercised.
  `"non_converged"` is excluded from Tier 1 by `filter_finemap_summary.py` (Wave 5).
- `L_saturated == TRUE` indicates all L=10 SuSiE effects were retained; a supplementary L=20 rerun
  is available per policy (flag only; no inline refit).
- `is_complex_region == TRUE` indicates the region is one of 4 pre-specified complex regions
  (HLA_6p21, APOE_19q13, 9p21_CDKN2A, SLC2A9_urate) per `config/susie_policy.yaml`.
- D4 columns (`n_CS_macor_*`) report the number of credible sets surviving at each
  `min_abs_corr` threshold. Monotonically non-increasing by construction.
- D5 diagnostic plots (regional association + LD heatmap) are generated on-demand from
  cached `.fit.rds` files, per CONTEXT.md G5 design. Not inlined in this dashboard.
